Microplatforms: Product platforms as code

### 2018-03-14 paper.tex
Minor edits and fixes

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the platform it runs on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See \ref{fig:concerns}).

@@

There has been a clear trend over the past two decades of automating as much of the above activities and treating them as another part of the product itself. This is the core proposition of cloud services and a key concept of the DevOps movement~\cite{DevOpsCu34:online}.

@@

In section \ref{sec:benefits} we outline the goals a good product platform should aim to support. In section \ref{sec:cost} the current approach of using a shared platform is discussed and an alternative (Microplatforms) is suggested in section \ref{sec:microplatforms}, supported by the concept of managing all aspects of it as source code in section \ref{sec:ascode}. Finally, we describe a reference architecture of an example Microplatform in Appendix \ref{sec:architecture}.

@@

The former extreme is very costly to scale and requires a lot of engineering effort to support, while the latter limits the product team’s ability to deliver exactly what is needed by the customer, as the team has to work around decisions made by someone else (and for someone else). This is an example of the 80/20 rule where a shared platform will get you 80% of the way very easily and the rest becomes a significant challenge, which just ends up frustrating the efforts of the team and the business.

@@

With increasing level of automation, the number of engineers necessary to support a product grows significantly slower with growth in the number of customers using that product, compared to more manual solutions [ref to google SRE]. As a corollary, with automation, the customer value delivered grows faster with the number of engineers (as more engineers add even more automation, boosting productivity even more), while without it, adding more engineers relatively quickly brings diminishing returns (figure \ref{fig:engineers-needed}). Automation enables relatively small teams of engineers to run products relying on a complex infrastructure whilst supporting millions of customers [ref WhatsApp].

Looking at the relationship another way, customer value delivered that leverages automation grows superlinearly with the number of people supporting the application, while without automation, it grows linearly at best, more likely sub-linearly due to coordination costs incurred from the lack of autonomy. This is one of the reasons start-up companies outperform large corporations: their resource constraints mean they are forced to invest in automation and smart solutions over adding more people, which in turn results in a much more efficient system. When we employ automation, we move the engineering effort higher up the value chain towards customer facing features.

@@

As an added benefit, the service orchestration is immutable. In order to change the configuration or topology of the microservices, the code describing it is changed and the orchestration redeployed. The cluster orchestration system then reconciles the differences so that only the parts that are different actually change. Repeated deployments and updates are idempotent - if there is no difference, nothing changes. This declarative approach means we can logically treat the whole service orchestration as a single unit for deployment and let the automation handle the details. We show the specifics of making sure this property holds in Appendix A.

In Fig 1 we show 48 cross-cutting concerns, identified by Sean Treadway~\cite{Treadway17:online}, that need addressing for each microservice. None of them are directly related to adding customer value. With microservices architecture, we have made a rod for our own back: developers should be able to write code that directly adds customer value and they should not need to worry about any of the concerns listed. These can be addressed by employing a service mesh (such as Istio []). It allows us to declaratively specify \textbf{policy as code} providing services such as intelligent routing and load balancing, traffic shaping and shifting, security policies, ingress and egress rules, circuit breaking [] and upstream call retries, rate limits, call tracing, telemetry and reporting. It is even possible to use a service mesh to inject faults into the system to test reliability features that may have been overlooked. All of this configuration is declarative and stored as code alongside the orchestration configuration.

@@

Immutability is typically quite easily attained by content addressing - referring to any data by a cryptographic hash of the content itself. This concept appears a few times in our approach: code stored in git is referred to by the sha1 hash of the contents of a given revision [], Docker images and layers can be immutably referred to by a sha256 hash of their content [].

@@

All of the above - automation, immutability, idempotence and autonomy are cases of simplicity, simple being distinctly different from easy [simple made easy]. By simple, we mean having a minimum of structure, little coupling and entanglement, avoiding N:M relationships between entities. Immutability is also a case of decoupling - mutable state is one of the hardest to understand couplings - a coupling through time.

>>>>

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the platform it runs on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See figure \ref{fig:concerns}).

@@

There has been a clear trend over the past two decades of automating as much of the above activities as possible and treating them as another part of the product itself. This is the core proposition of cloud services and a key concept of the DevOps movement~\cite{DevOpsCu34:online}.

@@

In section \ref{sec:benefits} we outline the goals a good product platform should aim to support. In section \ref{sec:cost} the current approach of using a shared platform is discussed and an alternative (Microplatforms) is suggested in section \ref{sec:microplatforms}, supported by the concept of managing all aspects of it as source code in section \ref{sec:ascode}. Finally, we describe a reference architecture of an example Microplatform in Appendix~\ref{sec:architecture}.

@@

The former extreme is very costly to scale and requires a lot of engineering effort to support, while the latter limits the product team’s ability to deliver exactly what is needed by the customer, as the team has to work around decisions made by someone else (and for someone else). This is an example of the 80/20 rule where a shared platform will get you 80\% of the way very easily and the rest becomes a significant challenge, which just ends up frustrating the efforts of the team and the business.

@@

With increasing level of automation, the number of engineers necessary to support a product grows significantly slower with growth in the number of customers using that product, compared to more manual solutions [ref to google SRE] (figure \ref{fig:engineers-needed}). As a corollary, with automation, the customer value delivered grows faster with the number of engineers (as more engineers add even more automation, boosting productivity even more), while without it, adding more engineers relatively quickly brings diminishing returns (figure \ref{fig:value-engineer}). Automation enables relatively small teams of engineers to run products relying on a complex infrastructure whilst supporting millions of customers [ref WhatsApp]This is one of the reasons start-up companies outperform large corporations: their resource constraints mean they are forced to invest in automation and smart solutions over adding more people, which in turn results in a much more efficient system. When we employ automation, we move the engineering effort higher up the value chain towards customer facing features.

@@

As an added benefit, the service orchestration is immutable. In order to change the configuration or topology of the microservices, the code describing it is changed and the orchestration redeployed. The cluster orchestration system then reconciles the differences so that only the parts that are different actually change. Repeated deployments and updates are idempotent - if there is no difference, nothing changes. This declarative approach means we can logically treat the whole service orchestration as a single unit for deployment and let the automation handle the details. We show the specifics of making sure this property holds in Appendix~\ref{sec:architecture}.

In figure \ref{fig:concerns} we show 48 cross-cutting concerns, identified by Sean Treadway~\cite{Treadway17:online}, that need addressing for each microservice. None of them are directly related to adding customer value. With microservices architecture, we have made a rod for our own back: developers should be able to write code that directly adds customer value and they should not need to worry about any of the concerns listed. These can be addressed by employing a service mesh (such as Istio []). It allows us to declaratively specify \textbf{policy as code} providing services such as intelligent routing and load balancing, traffic shaping and shifting, security policies, ingress and egress rules, circuit breaking [] and upstream call retries, rate limits, call tracing, telemetry and reporting. It is even possible to use a service mesh to inject faults into the system to test reliability features that may have been overlooked. All of this configuration is declarative and stored as code alongside the orchestration configuration.

@@

Immutability is typically quite easily attained by content addressing - referring to any data by a cryptographic hash of the content itself. This concept appears a few times in our approach: code stored in git is referred to by the sha1 hash of the contents of a given revision [], Docker images and layers can be immutably referred to by a \texttt{sha256} hash of their content [].

@@

All of the above - automation, immutability, idempotence and autonomy are cases of simplicity, \textit{simple} being distinctly different from \textit{easy} [simple made easy]. By simple, we mean having a minimum of structure, little coupling and entanglement, avoiding N:M relationships between entities. Immutability is also a case of decoupling - mutable state is one of the hardest to understand couplings - a coupling through time.
